The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-07-19 paper.qmd
tried to get bottom align to work (didnt) and added some decision bit

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-ibrekk
#| fig-cap: "A replication of the the uncertainty visualisations shown by @Ibrekk1987 in one of the earliest uncertainty visualisation experiments. This early experiment is a good example of many of the issues that are still common in uncertainty visualisation today. For example, the '95% confidence interval' is more accurately a '95% prediction interval'. Additionally graphics that depict different mathematical objects that are also different on their visual components are compared because of a percieved relation to uncertainty. So visualisations of the mean, PDF and CDF, are all discussed as though they all contain the relevant statistical information. The axis also have different scales, and visualisation methods that are now unpopular for displaying proportions, such as a pie chart, are used."
#| fig-subcap: 
#|   - "Picture 1"
#|   - "Picture 2"
#|   - "Picture 3"
#|   - "Picture 4"
#|   - "Picture 5"
#|   - "Picture 6"
#|   - "Picture 7"
#|   - "Picture 8"
#|   - "Picture 9"
#| layout-valign: bottom
#| layout-ncol: 3

@@

That is to say, despite the wealth of reviews, the field of uncertainty visualisation remains a heap of stones. This review attempts to address this issue by offering a novel perspective on the uncertainty visualisation problem, and hopefully laying the foundations on which we can build a house.

This review is broken into several parts that each reflect a different approach to uncertainty visualisation. First we look at graphics that ignore uncertainty entirely and discuss why uncertainty should be included at all. Second, we look at methods that consider uncertainty to be just another variable and discuss the characteristics of uncertainty that make it a unique visualisation problem. Third, we look at methods that explicitly combine our estimate and its uncertainty and discuss if the visualisations created by these transformations are still "uncertainty visualisations". Fourth, we will discuss methods that implicitly include uncertainty by depicting a sample or original data in place of an estimate. Finally, we discuss how uncertainty visualisations can be effectively evaluated. When discussing each of these methods, we consider the *purpose* of uncertainty visualisation and comment on how effective each visualisation is at fulfilling that purpose.

@@

# Ignoring uncertainty
Before we look at different cases where uncertainty is included, it is beneficial for us to look at cases where it is not. After all, why include uncertainty at all?

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-choropleth
#| fig-cap: "Two choropleth maps that depicts the counties of Iowa where each country coloured acording to a simulated average temperature. Both maps depict a spatial trend, where counties closer to the center of the map are hotter than counties on the edge of the map. This trend is only technically there in the low variance condition, i.e. if we performed a hypothesis test we would conclude the trend exists. The high variance condition displays a spatial trend that could simply be the result of noise, which means the plot is displaying a false conclusion."
#| fig-subcap: 
#|   - "High Variance Data"
#|   - "Low Variance Data"
#|   - ""
#| layout-ncol: 3
#| layout-valign: "bottom"

@@

The two choropleth maps appearing to be identical in @fig-choropleth highlights the need for uncertainty visualisations. Uncertainty visualisation is is required for transparency and this sentiment has been repeated.  Some authors suggest uncertainty is important to include as it communicates the legitimacy (or illegitimacy) of the conclusion drawn from visual inference [@Correll2014; @Kale2018; @Griethe2006]. Some authors have said that uncertainty should be included to degree of confidence or trust in the data [@Boukhelifa2012; @Zhao2023]. Some authors directly connect uncertainty visualisation to hypothesis testing as it ensures the "validity" of a statement [@Hullman2020a; @Griethe2006], but allows for a proportional level of trust that is more detailed than the binary results of a hypothesis test [@Correll2014; @Correll2018]. Some authors even go so far as to claim that failing to include uncertainty is akin to fraud or lying [@Hullman2020a; @Manski2020].

@@

As @fig-choropleth showed, visualisations that are unconcerned with uncertainty have no issue showing justified signals, but struggle with the display of unjustified signals. Therefore, we call this approach "signal-supression" as it primarily differentiates itself from from a typical visualisation approach through criteria (2), that is, an uncertainty visualisation should prevent us from drawing unjustified conclusions. We will investigate several approaches for uncertainty visualisation in the following sections and in doing so, discuss how successful each approach is in fulfilling the signal-suppression criteria.

@@

## Bivariate Map
## Signal Supression

@@

## Can we have uncertainty without signal supression?
Some authors do not describe uncertainty as a means of transparency, but rather as a variable of importance in of itself [@Blenkinsop2000].

It is not entirely true that uncertainty is only important for transparency, however it is not clear that the "uncertainty" provided in other cases is uncertainty at all. Some authors claim the purpose of uncertainty is not transparency, but rather important information for effective decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017]. While this can be true, the uncertainty provided for this purpose is *not* the same as the uncertainty provided for transparency. It arguably may not be uncertainty at all. This is easiest to understand with an example.

Let's say my weather prediction app only provides me with the predicted daily rainfall when I am trying to decide if I should bring an umbrella. While my decision will be improved with the inclusion of uncertainty information, that is not because of a special quality of uncertainty, but rather because the uncertainty information can be used to calculate the true statistic of interest, the probability that it will rain at all. If I use uncertainty to improve decision making, it is not *true* uncertainty, rather it is supplementary material that is required to correct for an insufficient statistic. I was provided with was not the statistic required to make the decison.

# Explicitly Combining Uncertainty and Signal
## Using Statistics
### Exceedance Probability Map
### What is an uncertainty visualisation?
## Using Aesthetics
### Value Supressing Uncertainty Palettes
### There is no such thing as a test for everything

>>>>

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-ibrekk
#| fig-cap: "A replication of the the uncertainty visualisations shown by @Ibrekk1987 in one of the earliest uncertainty visualisation experiments. This early experiment is a good example of many of the issues that are still common in uncertainty visualisation today. For example, the '95% confidence interval' is more accurately a '95% prediction interval'. Additionally graphics that depict different mathematical objects that are also different on their visual components are compared because of a percieved relation to uncertainty. So visualisations of the mean, PDF and CDF, are all discussed as though they all contain the relevant statistical information. The axis also have different scales, and visualisation methods that are now unpopular for displaying proportions, such as a pie chart, are used."
#| fig-subcap: 
#|   - "Picture 1"
#|   - "Picture 2"
#|   - "Picture 3"
#|   - "Picture 4"
#|   - "Picture 5"
#|   - "Picture 6"
#|   - "Picture 7"
#|   - "Picture 8"
#|   - "Picture 9"
#| layout-ncol: 3
#| fig-cap-location: "margin"

@@

That is to say, despite the wealth of reviews, the field of uncertainty visualisation remains a heap of stones. This review attempts to address this issue by offering a novel perspective on the uncertainty visualisation problem, and hopefully laying the foundations on which we can build a house. 
This review is broken into several parts that each reflect a different approach to uncertainty visualisation. First we look at graphics that ignore uncertainty entirely and discuss why uncertainty should be included at all. Second, we look at methods that consider uncertainty to be just another variable and discuss the characteristics of uncertainty that make it a unique visualisation problem. Third, we look at methods that explicitly combine our estimate and its uncertainty and discuss if the visualisations created by these transformations are still "uncertainty visualisations". Fourth, we will discuss methods that implicitly include uncertainty by depicting a sample or original data in place of an estimate. Finally, we discuss how uncertainty visualisations can be effectively evaluated. When discussing each of these methods, we consider the *purpose* of uncertainty visualisation and comment on how effective each visualisation is at fulfilling that purpose.

@@

# Ignoring uncertainty
A good place to start might be at deceptively straight forward question, why should we include uncertainty at all?

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-choropleth
#| fig-cap: "Two choropleth maps that depict the counties of Iowa where each country coloured acording to a simulated average temperature. Both maps depict a spatial trend, where counties closer to the center of the map are hotter than counties on the edge of the map. This trend is only technically there in the low variance condition, i.e. if we performed a hypothesis test we would conclude the trend exists. The high variance condition displays a spatial trend that could simply be the result of noise, which means the plot is displaying a false conclusion."
#| fig-subcap: 
#|   - "High Variance Data"
#|   - "Low Variance Data"
#|   - "Choropleth Palette"
#| layout-ncol: 3
#| layout-valign: "bottom"

@@

## Signal-supression
The two choropleth maps appearing to be identical in @fig-choropleth highlights the need for uncertainty visualisations. Uncertainty visualisation is is required for transparency and this sentiment has been repeated.  Some authors suggest uncertainty is important to include as it communicates the legitimacy (or illegitimacy) of the conclusion drawn from visual inference [@Correll2014; @Kale2018; @Griethe2006]. Some authors have said that uncertainty should be included to degree of confidence or trust in the data [@Boukhelifa2012; @Zhao2023]. Some authors directly connect uncertainty visualisation to hypothesis testing as it ensures the "validity" of a statement [@Hullman2020a; @Griethe2006], but allows for a proportional level of trust that is more detailed than the binary results of a hypothesis test [@Correll2014; @Correll2018]. Some authors even go so far as to claim that failing to include uncertainty is akin to fraud or lying [@Hullman2020a; @Manski2020].

@@

We can refer to the we draw from a graphic to be its "signal" and the variance that makes this signal harder to identify as the "noise". As @fig-choropleth showed, visualisations that are unconcerned with uncertainty have no issue showing justified signals, but struggle with the display of unjustified signals. Therefore, we coin this approach to uncertainty visualisation as "signal-supression" since it primarily differentiates itself from from the "noiseless" visualisation approach through criteria (2). That is, the main difference between an uncertainty visualisation and a "normal" visualisation, is that an uncertainty visualisation should prevent us from drawing unjustified conclusions.

## Decision making does not need uncertainty visualisation
Of course, uncertainty visualisation is not only motivated by signal-suppression. Some authors claim the purpose of uncertainty is not transparency, but rather important information for effective decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017]. While uncertainty can provide useful information in decision making, that information is actually signal that is required due to incomplete information. This is easiest to understand with an example.

Imagine you are trying to decide if you want to bring an umbrella with you to work. An umbrella is annoying to bring and you only want to bring it if there is a 10% chance of rain (or higher). Unfortunately, your weather prediction app only provides you with the predicted daily rainfall. Therefore, your decision will be improved with the inclusion of uncertainty, *not* because uncertainty is important for your decision, but because it gives you the tools required to calculate the *actual* statistic you are basing your decision on.

While my decision will be improved with the inclusion of uncertainty information, that is not because of a special quality of uncertainty, but rather because the uncertainty information can be used to calculate the true statistic of interest, the probability that it will rain at all. If I use uncertainty to improve decision making, it is not *true* uncertainty, rather it is supplementary material that is required because the original information was not sufficient. In this sense uncertainty is no different from any other variable. If someone is asked to calculate a persons BMI and they are only given their height, of *course* their calculation will be improved if you provide their weight too.

We can actually prove this result is true mathematically. Let $A$ be a rational agent that needs to make a decision and let $U(p)$ be the utility function of agent $A$ that depends on the random variable, $p$ where $p$ is a probability. Additionally $p$ can be calculated using a density function $p=f(\mu, \sigma)$

The ma

The best "uncertainty visualisation" in decision making experiments is reliably the visualisation that makes the calculation of

For these reasons decision making as a motivation for uncertainty visualisation will not be the focus of this review, as the uncertainty in this case is not "noise" and does not require any special visualisation.

### Exceedance Probability Map
For these reasons, if the statistic required to make the decision is visualised, there is no reason to include uncertainty on *that* graphic.

the uncertainty provided for this purpose is *not* the same as the uncertainty provided for transparency.

We will investigate several approaches for uncertainty visualisation in the following sections and in doing so, discuss how successful each approach is in fulfilling the signal-suppression criteria.

@@

## Bivariate Map
## Uncertainty is not just another variable

Some authors do not describe uncertainty as a means of transparency, but rather as a variable of importance in of itself [@Blenkinsop2000].

@@

# Explicitly Combining Uncertainty and Signal
## Using Statistics
### What is an uncertainty visualisation?
## Using Aesthetics
### Value Supressing Uncertainty Palettes
### There is no such thing as a test for everything
